---
title: "Auxiliary Trait Selection with Coheterogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Auxiliary Trait Selection with Coheterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(IBMR)
```

# Overview

This vignette describes the first step in the `IBMR` workflow: selecting an
auxiliary outcome trait for a given exposure and primary outcome.

The central motivation is that closely related outcomes may share overlapping
valid instruments, or may reflect similar mechanisms of instrument invalidity,
with respect to a common exposure. If this shared structure can be identified,
an auxiliary trait may be used to improve downstream MR estimation and outlier
detection.

`IBMR` operationalizes this idea through `coheterogeneity_Q()`, which computes
pairwise coheterogeneity between outcome traits using SNP-level summary
statistics.

# The Data Structure

At minimum, the coheterogeneity analysis requires:

- `BetaXG`: SNP-exposure associations
- `seBetaXG`: standard errors for `BetaXG`
- `BetaYG_matrix`: a matrix of SNP-outcome associations
- `seBetaYG_matrix`: a matrix of standard errors corresponding to
  `BetaYG_matrix`

Each row corresponds to a SNP, and the rows must be aligned across all vectors
and matrices.

In the common screening setting, the columns of `BetaYG_matrix` are:

- one primary outcome
- one or more candidate auxiliary outcomes

# Toy Example Data

The package includes a toy dataset called `toy_ibmr_example`.

```{r}
data("toy_ibmr_example")

names(toy_ibmr_example)
toy_ibmr_example$primary_trait
toy_ibmr_example$candidate_auxiliaries
```

The object contains:

- one exposure vector
- one primary outcome
- two candidate auxiliary traits
- a small formatted data frame for `IBPRESSO()`
- metadata describing the simulation setup

The simulation is deliberately simple. It was constructed so that
`aux_trait_1` shares more latent structure with the primary outcome than
`aux_trait_2`, making it the more natural auxiliary trait in this example.

```{r}
str(toy_ibmr_example$simulation_setup)
```

# Running the Coheterogeneity Analysis

We begin by extracting the summary-statistics objects.

```{r}
BetaXG <- toy_ibmr_example$BetaXG
seBetaXG <- toy_ibmr_example$seBetaXG
BetaYG_matrix <- toy_ibmr_example$BetaYG_matrix
seBetaYG_matrix <- toy_ibmr_example$seBetaYG_matrix

colnames(BetaYG_matrix)
```

We now compute coheterogeneity across the primary outcome and the candidate
auxiliary traits.

```{r}
cohet_res <- coheterogeneity_Q(
  BetaXG = BetaXG,
  BetaYG_matrix = BetaYG_matrix,
  seBetaXG = seBetaXG,
  seBetaYG_matrix = seBetaYG_matrix,
  F_min = 5,
  min_K_pair = 20
)
```

The principal outputs are:

- `rho`: pairwise coheterogeneity estimates
- `p_value`: pairwise significance levels
- `flag`: diagnostic status for each pair
- `K`: number of SNPs contributing to each pairwise calculation

```{r}
round(cohet_res$rho, 3)
round(cohet_res$p_value, 3)
cohet_res$flag
cohet_res$K
```

# Interpreting the Output

The diagonal entries of `rho` are equal to 1 by definition. The quantities of
primary interest are the off-diagonal entries involving the primary outcome.

```{r}
primary_name <- toy_ibmr_example$primary_trait
cohet_res$rho[primary_name, ]
cohet_res$p_value[primary_name, ]
cohet_res$flag[primary_name, ]
```

In this toy example, `aux_trait_1` should usually exhibit stronger
coheterogeneity with the primary outcome than `aux_trait_2`.

A practical interpretation framework is:

- larger `abs(rho)` indicates a stronger shared heterogeneity pattern
- smaller `p_value` provides stronger evidence against no coheterogeneity
- `flag` should be checked to ensure the pairwise estimate is usable

The most important diagnostic flags are:

- `"OK"`: the pairwise comparison was successfully estimated
- `"tau0"`: the corrected heterogeneity estimate was not identifiable
- `"se NA"`: the coheterogeneity estimate was computed, but its standard error
  could not be reliably estimated

# Ranking Candidate Auxiliary Traits

The screening step is usually carried out by ranking candidate auxiliaries for
the primary outcome.

```{r}
ranking <- data.frame(
  aux_trait = colnames(cohet_res$rho),
  rho = cohet_res$rho[primary_name, ],
  p_value = cohet_res$p_value[primary_name, ],
  flag = cohet_res$flag[primary_name, ],
  stringsAsFactors = FALSE
)

ranking <- subset(ranking, aux_trait != primary_name)
ranking$abs_rho <- abs(ranking$rho)
ranking <- ranking[order(-ranking$abs_rho), ]
ranking
```

A practical selection rule is:

1. discard pairs with unusable flags
2. prioritize pairs with finite and informative `p_value`
3. among the remaining candidates, select the trait with the largest `abs(rho)`

In this toy dataset, the intended selected auxiliary trait is:

```{r}
toy_ibmr_example$recommended_auxiliary
```

# Notes on Guardrails

`coheterogeneity_Q()` includes several guardrails to stabilize estimation:

- optional SNP restriction through `SNP_keep`
- weak-instrument filtering through `F_min`
- small-exposure filtering through `bx_min`
- optional winsorization through `winsor_theta`
- pair-level diagnostic flags

These are especially useful in large-scale screening settings where some traits
may have sparse or unstable SNP-level signal.

# Next Step

Once an auxiliary trait has been selected, it can be carried into downstream
instrument-borrowing analysis using:

- `IBMODE()` for multidimensional mode-based estimation
- `IBPRESSO()` for outlier detection and correction with an auxiliary trait

The end-to-end downstream workflow is illustrated in the companion vignette:

- `instrument-borrowing-workflow`
