---
title: "Instrument Borrowing Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Instrument Borrowing Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(IBMR)
```

# Overview

This vignette illustrates the full `IBMR` workflow:

1. load SNP-level summary statistics
2. screen candidate auxiliary traits using coheterogeneity
3. select the most informative auxiliary trait
4. use the selected auxiliary trait in `IBMODE()`
5. use the same auxiliary trait in `IBPRESSO()`

The purpose is to demonstrate how auxiliary-trait selection and downstream
instrument borrowing fit together in a single analysis pipeline.

# Load the Toy Example

```{r}
data("toy_ibmr_example")

BetaXG <- toy_ibmr_example$BetaXG
seBetaXG <- toy_ibmr_example$seBetaXG
BetaYG_matrix <- toy_ibmr_example$BetaYG_matrix
seBetaYG_matrix <- toy_ibmr_example$seBetaYG_matrix
primary_name <- toy_ibmr_example$primary_trait
candidate_aux <- toy_ibmr_example$candidate_auxiliaries
```

The toy example contains one primary outcome and two candidate auxiliary
outcomes. The data were simulated so that `aux_trait_1` is usually the more
informative auxiliary trait for the primary outcome.

# Step 1: Screen Auxiliary Traits with Coheterogeneity

```{r}
cohet_res <- coheterogeneity_Q(
  BetaXG = BetaXG,
  BetaYG_matrix = BetaYG_matrix,
  seBetaXG = seBetaXG,
  seBetaYG_matrix = seBetaYG_matrix,
  F_min = 5,
  min_K_pair = 20
)

round(cohet_res$rho, 3)
round(cohet_res$p_value, 3)
cohet_res$flag
```

We focus on the row corresponding to the primary outcome.

```{r}
rho_primary <- cohet_res$rho[primary_name, ]
p_primary <- cohet_res$p_value[primary_name, ]
flag_primary <- cohet_res$flag[primary_name, ]

rho_primary
p_primary
flag_primary
```

# Step 2: Rank Candidate Auxiliary Traits

```{r}
ranking <- data.frame(
  aux_trait = candidate_aux,
  rho = rho_primary[candidate_aux],
  p_value = p_primary[candidate_aux],
  flag = flag_primary[candidate_aux],
  stringsAsFactors = FALSE
)

ranking$abs_rho <- abs(ranking$rho)
ranking <- ranking[order(-ranking$abs_rho), ]
ranking
```

The selected auxiliary trait can be taken as the candidate with the largest
absolute coheterogeneity among those with acceptable diagnostic flags.

```{r}
chosen_aux <- ranking$aux_trait[1]
chosen_aux
```

For the packaged toy example, the intended selected auxiliary trait is:

```{r}
toy_ibmr_example$recommended_auxiliary
```

# Step 3: Run `IBMODE()`

Once the auxiliary trait has been selected, we subset the outcome matrices to
the primary outcome and the chosen auxiliary trait.

```{r}
BetaYG_mode <- BetaYG_matrix[, c(primary_name, chosen_aux), drop = FALSE]
seBetaYG_mode <- seBetaYG_matrix[, c(primary_name, chosen_aux), drop = FALSE]
```

We then run `IBMODE()`.

```{r, eval = FALSE}
ibmode_res <- IBMODE(
  BetaXG = BetaXG,
  BetaYG_matrix = BetaYG_mode,
  seBetaXG = seBetaXG,
  seBetaYG_matrix = seBetaYG_mode,
  phi = c(1, 0.5),
  n_boot = 200
)

ibmode_res
```

This returns joint mode-based estimates for the primary and auxiliary outcomes.
In routine analyses, the number of bootstrap replicates should be chosen to
balance computational cost and stability.

# Step 4: Run `IBPRESSO()`

`IBPRESSO()` expects a data frame rather than matrices. The toy dataset
includes a ready-made example for the recommended auxiliary trait.

```{r}
dat_ibpresso <- toy_ibmr_example$dat_ibpresso_aux1
head(dat_ibpresso)
```

The analysis can be run as follows.

```{r, eval = FALSE}
ibpresso_res <- IBPRESSO(
  BetaOutcome = "beta_primary",
  BetaExposure = "beta_exposure",
  BetaAux = "beta_aux",
  SdOutcome = "se_primary",
  SdExposure = "se_exposure",
  SdAux = "se_aux",
  data = dat_ibpresso,
  OUTLIERtest = TRUE,
  DISTORTIONtest = FALSE,
  NbDistribution = 200,
  seed = 123
)

ibpresso_res
```

This analysis uses the selected auxiliary trait to improve outlier detection in
the joint primary-auxiliary residual space.

# Interpreting the Workflow

The key logic of the instrument-borrowing approach is:

1. identify a related auxiliary trait with meaningful coheterogeneity
2. preserve the primary outcome as the principal target of inference
3. use the auxiliary trait only to stabilize downstream robust MR analysis

The auxiliary trait is therefore not an alternative endpoint. Instead, it is an
informative companion trait used to strengthen inference for the primary
outcome.

# Practical Recommendations

- Ensure that SNP order is aligned across all inputs.
- Harmonize alleles before constructing summary-statistics matrices.
- Examine `flag` and `K` from `coheterogeneity_Q()` before selecting an
  auxiliary trait.
- Use more than one candidate auxiliary trait whenever scientifically
  reasonable, since screening is most useful when several plausible auxiliaries
  are available.
- Increase `n_boot` in `IBMODE()` and `NbDistribution` in `IBPRESSO()` for
  final analyses.

# Summary

`IBMR` supports a two-stage strategy:

- first, identify a suitable auxiliary trait using coheterogeneity
- second, use that auxiliary trait to improve robust MR estimation and outlier
  detection

This framework is particularly useful when related outcome traits are expected
to share informative instrument-validity structure relative to a common
exposure.
